The Concurrent Calculi Formalisation Benchmark

### 2024-02-22 short-paper/short-paper.tex
Minor fixes.
Delete old TeX comments.

% - Introduction
%   - POPLMark accelerated progress in mechanisation tools for PL
%     - POPLMark Reloaded tried to do the same for logical relations
%     - But they were deliberately limited in scope to not include process calculi
%   - We propose a new collection of benchmark problems for process calculi
%     - Why not concurrency in general? (after this point, only mention process calculi)
%     - Why is there a problem with process calculus mechanisations?
%     - Our experience says there is, and we will cite other experiences below
%   - We have noticed three key issues
%     - They are not covered by POPLMark
%     - Techniques are less settled than for binders
%     - Others also mention them as problems: cite and explain references briefly for each challenge topic
%   - Our goals are to determine the state of the art, develop best practices and tutorials, and identify improvements to proof assistant technology that would materially help the situation
\section{Introduction}
The POPLMark challenge~\cite{POPLMark}
played a pivotal role in advancing the field of proof assistants,
libraries, and best practices for programming language
mechanisation. By providing a shared framework for systematically
evaluating mechanisation techniques, it catalysed a significant
shift towards publications that include mechanised proofs within the programming language research community.
POPLMark Reloaded~\cite{POPLMarkReloaded} introduced a similar programme targeting proofs using logical relations.
These initiatives had a narrow focus, but their authors recognised the importance of addressing topics such as coinduction and linearity in the future.

@@

\subsection{Scope extrusion}
This challenge revolves around the mechanisation of % explicit
scope
extrusion, by which a process can send restricted names to another
process, as long as the restriction can safely be extruded to include
the receiving process.  The setting for this challenge is a
``classic'' untyped \( \pi \)-calculus, where (unlike the calculi in
the other challenges) names can be sent and received, and bound by
input constructs.  We define two different semantics for our system:
\begin{enumerate}
\item A reduction system: this avoids explicit reasoning about scope
  extrusion by using instead structural congruence, allowing
  implementors to explore alternative ways to encode the latter (\eg
  via process contexts or via compatible refinement);
\item An (early) labelled transition system.
\end{enumerate}
The objective of this
challenge is to prove that the two semantics are equivalent up to
structural congruence.

@@

% Higher-order abstract syntax (HOAS) has been extensively and
% successfully used~\cite{Despeyroux2000,Honsell2001,Tiu2010}, since in
% its weak form it is compatible with mainstream inductive proof
% assistants. While this approach disposes of most of the technicalities
% when encoding typing systems or operational semantics, it requires
% extensions when dealing with notions such as bisimulation, either in the
% form of additional axioms~\cite{Honsell2001}, or of dedicated proofs
% assistants such as Abella, which features a specific
% quantifier~\cite{GacekMN11} to handle properties of names.

@@

Process calculi typically include constructs that allow processes to
have infinite behaviours.  Coinduction serves as a fundamental method
for the definition and analysis of infinite objects, enabling the
examination of their behaviours.
% Coinduction is a fundamental technique for
% defining and reasoning about infinite objects, and it can thus be used
% to reason about these behaviours.

@@

While many proof assistants support coinductive techniques, they do so
through subtly different formalisms, enabling various
approaches. Additionally, some systems offer multiple abstractions for
utilizing coinduction. Examples are Agda's musical notation versus  co-patterns and sized types~\cite{Abel2013}. In Coq, guarded recursion versus a refined fixed point approach as supported by the parameterised coinduction~\cite{Hur2013}
and interaction tree~\cite{Xia2019} libraries.

When reasoning over bisimilarity many authors rely on the native
 coinduction offered by the chosen proof assistants
\cite{Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, others prefer a
more ``set-theoretic'' approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}. Some
use both and establish an internal adequacy~\cite{Honsell2001}. Few
extend the proof assistant foundations to allow \eg reasoning about
bisimilarity up-to~\cite{ChaudhuriCM15}.
% While many proof assistants support
% coinductive techniques,   but they support subtly different formalisms that enable different techniques, and some systems provide more than one abstraction to use coinduction.
% We thus see a variety of different formalisation choices for
% bisimulations, sometimes ad-hoc for the specific result.

% \todo{Reorganize the rest of this section to be more coherent}
% Isabelle, Coq, and Abella seamlessly support coinductive definitions and reasoning.
% Agda and Beluga support it through the convenient abstraction of co-patterns~\cite{Abel2013}.
% Newer entrants to the field like the Lean proof assistant
% are working on adding a notion of coinduction to their system~\cite{Avigad2019}.\alb{I kinda object to ``seamlessly''. In Coq it's a mess, in Abella very very limited, sized types in Agda are shaky. I guess Isabelle is better}
%
% A good example of this are the parameterised coinduction~\cite{Hur2013}
% and interaction tree~\cite{Xia2019} libraries for Coq.

% E.g., \cite{Castro-Perez2021} notes that the most challenging
% part of their mechanisation of multiparty session types
% was the convoluted definition of infinite behaviours;
% \cite{Pohjola2022} points out that the main difficulty
% of the proofs concerning bisimulation lies
% in dealing with existential quantifiers; and
% \cite{Bengtson2016} wonders whether parametrised
% coinduction as developed in Coq \cite{Hur2013} would
% have helped their Isabelle mechanisation.

% \lore{Should we mention that Isabelle offers
%   the command \texttt{coinduction\_upto}? Is it related?}

% While many authors rely on the native support for coinduction offered
% by proof assistants
% \cite{Honsell2001,Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, many
% others prefer a more ``set-theoretic'' approach
% \cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}.
%\alb{Actually, Honsell gives up with guarded coinduction and switches to Park's half way}

>>>>

\section{Introduction}
The POPLMark challenge~\cite{POPLMark}
played a pivotal role in advancing the field of proof assistants,
libraries, and best practices for programming language
mechanisation. By providing a shared framework for systematically
evaluating mechanisation techniques, it catalysed a significant
shift towards publications that include mechanised proofs within the programming language research community.
POPLMark Reloaded~\cite{POPLMarkReloaded} introduced a similar programme targeting proofs using logical relations.
These initiatives had a narrow focus, but their authors recognised the importance of addressing topics such as coinduction and linearity in the future.

@@

\subsection{Scope extrusion}
This challenge revolves around the mechanisation of scope
extrusion, by which a process can send restricted names to another
process, as long as the restriction can safely be extruded to include
the receiving process.  The setting for this challenge is a
``classic'' untyped \( \pi \)-calculus, where (unlike the calculi in
the other challenges) names can be sent and received, and bound by
input constructs.  We define two different semantics for our system:
\begin{enumerate}
\item A reduction system: this avoids explicit reasoning about scope
  extrusion by using instead structural congruence, allowing
  implementors to explore alternative ways to encode the latter (\eg
  via process contexts or via compatible refinement);
\item An (early) labelled transition system.
\end{enumerate}
The objective of this
challenge is to prove that the two semantics are equivalent up to
structural congruence.

@@

@@

Process calculi typically include constructs that allow processes to
have infinite behaviours.  Coinduction serves as a fundamental method
for the definition and analysis of infinite objects, enabling the
examination of their behaviours.

@@

While many proof assistants support coinductive techniques, they do so
through subtly different formalisms, enabling various
approaches. Additionally, some systems offer multiple abstractions for
utilising coinduction. Examples are Agda's musical notation versus  co-patterns and sized types~\cite{Abel2013}. In Coq, guarded recursion versus a refined fixed point approach as supported by the parameterised coinduction~\cite{Hur2013}
and interaction tree~\cite{Xia2019} libraries.

When reasoning over bisimilarity many authors rely on the native
 coinduction offered by the chosen proof assistants
\cite{Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, others prefer a
more ``set-theoretic'' approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}. Some
use both and establish an internal adequacy~\cite{Honsell2001}. Few
extend the proof assistant foundations to allow \eg reasoning about
bisimilarity up-to~\cite{ChaudhuriCM15}.
